---
title: "Benthic_community_clean"
output: html_notebook
author: Charlotte E Page
---


# NFI Benthic Cover data analysis and PCA


This is a script for data clean up and analysis of benthic cover data. Data collected from photoquadrats in Emily and Slaughter Bay, over two time points. 

TP1: Dec (2020)
TP2: April (2021)

Total of n = 12 transects per TP. n ~ 10 photoquads per transect (note: in some instances n is lower than 0, this was due to error in the field or in post-processing)

All hard coral and algal taxa were identified to genera. Where multiple growth forms existed within each genera, data has been originally binned into these growth forms (within genera). Growth form bins for each genera will be grouped to genera for quantification of cover. 

Script aims:  

A) Tidy data format (long)

B) Filter coral genera into categories based on cover.
      
C) PCA - Generate driver variables that can be used in modelling disease occurence and disease severity. 

Data download and tidy

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggthemes)
library(factoextra) 
library(tidymodels)

# Visualising PCA analysis results

cover <- read.csv(file= "NFI_cover_data.csv", header = TRUE)

str(cover)

```

### A) Tidy data format (long)


```{r}

# Transpose the data frame
t.cover <- cover %>% gather(Genera, Cover, Acanthastrea:Valonia)
str(t.cover)

# Generate some summary plots

# Summarise cover by Date, genera and transect
t.cover.freq <- t.cover %>% group_by(Date,Transect_ID,Genera) %>%
  summarise(freq = mean(Cover)) 

# Average across transects for plot 
cover.sum <- t.cover.freq %>% 
  group_by(Date, Genera) %>% 
  summarise(mean.freq =mean(freq),
  SD_freq = sd(freq),
  SE_freq = sd(freq)/sqrt(n()))

str(cover.sum)

ggplot(cover.sum, aes(x = Date, y = mean.freq, fill = Genera)) +
      geom_bar(position="fill", stat="identity") + scale_fill_viridis_d()

ggplot(t.cover.freq, aes(x = Genera, y = freq)) +
      geom_boxplot() +
      facet_wrap(~Date) +
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

Organise cover ID into categories 

```{r}

# Nested if else to create a new variable where genera are sorted into benthic category type.

# First list the values in the Genera column

# levels(t.cover.freq$Genera)

benthic.cats <- t.cover.freq %>%
  mutate(Category = 
      ifelse(Genera %in% c("Acanthastrea", "Acropora_br", "Acropora_pl", "Alveopora","Astrea","Euphyllia", "Montipora_enc","Montipora_fol","Paragoniastrea","Pectinia","Platygyra", "Plesiastrea","Pocillopora","Porites","Psammocora","Stylophora","Cyphastrea","Homophyllia.sp.","Lobophyllia","Turbinaria"),
             "Hard_coral",
      ifelse(Genera %in% c("Amansia","Amphiroa","Asparagopsis","Bryopsis","Caulerpa_cupre","Caulerpa_racemosa", "Chaetomorpha","Chlorodesmis.sp.","Codium.fragile","Colpomennia.sinuosa","Dasycladus","Dictyopteris","Dictyota","Halimeda","Halymenia","Homophyllia.sp", "Hormosira","Hydroclathrus","Hypnea","Laurencia","Padina","Peysonnelia","Plocamium","Sargassum","Ulva","Valonia","Tricleocarpa.fragilis"), 
             "Macroalgae",
      ifelse(Genera %in% c("Soft_coral","Anemone"), 
             "Soft_coral",
      ifelse(Genera %in% c("Sponge","Zooanthid","Benthic_invert"), 
             "Other_benthic_invert",
      ifelse(Genera %in% c("Black_cyano","Cyanofil","Red_cyano","Turf","Turf_sediment"),
               "Turf_mixed",
      ifelse(Genera %in% c("Sediment","Sand","Rubble","Dead_coral"),
             "Other_substrate", 
      ifelse(Genera %in% c("CCA","Rhodoliths"),
             "Coralline_algae", "NA"
             ))))))))

# Search for any NAs

filter(benthic.cats, Category == "NA")
# 0 rows!

```

Quick plot to visualise these benthic groups 

```{r}

relevel(benthic.cats$Date,"April")

ggplot(benthic.cats, aes(x = Category, y = freq, fill = Date)) +
  geom_boxplot()

ggplot(benthic.cats, aes(x = Date, y = freq, fill = Category)) +
      geom_bar(position="fill", stat="identity") +
      scale_fill_viridis_d()

```


### B) Filter coral genera into categories based on cover.


Find coral genera of interest (taxa < 10 % relative cover)


```{r}
# Caclulate the relative coral cover per group
# Convert Category variable to a factor 
str(benthic.cat)

benthic.cats <- as.data.frame(benthic.cats)

benthic.cats[,"Category"] <- as.factor(benthic.cats[,"Category"])


# Relevel Date so that December comes first 
benthic.cats[,"Date"] <- factor(benthic.cats[,"Date"], levels = c("Dec", "April"))

# Look at the levels
head(benthic.cats)

# Filter out Hard_coral
hard_coral <- benthic.cats %>% filter(Category == "Hard_coral")

# Calculate total hard coral cover (%)
hc_tran <- hard_coral %>% group_by(Date, Transect_ID) %>%
 mutate(t_hardcoral = sum(freq))

hc_trana <- hc_tran %>% mutate(rel_hardcoral = (freq/t_hardcoral)*100)

# Summarise data frame (per coral taxa, per time point)

coralrelsum <- hc_trana %>% group_by(Date,Genera) %>% 
  summarise(rel_sum = mean(rel_hardcoral, na.rm = TRUE),
            SD_freq = sd(rel_hardcoral),
            SE_freq = sd(rel_hardcoral/sqrt(n()))) 

# plot relative coral cover of genera over time
colours <- c("cyan2", "darkcyan")

plot <- ggplot(coralrelsum, aes(Genera, rel_sum, fill = Date)) +
 geom_col(position = "dodge",stat = "identity") +
 geom_errorbar(aes(ymin = rel_sum - SE_freq, ymax = rel_sum + SE_freq), 
               width=0.2, position = position_dodge(.9), color = "black") +
  scale_y_continuous(limits = c(0,40)) +
  #geom_label(stat = 'summary', fun.y=mean, aes(label = round(..y.., 1)), nudge_x =0.5 , hjust = 3) +
  xlab ("Taxa") +
  ylab ("Relative cover (%)") + 
  ggtitle("Relative cover hermatypic taxa") +
 theme_few() +
  scale_fill_manual(values = colours)

a <- plot + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + geom_hline(yintercept=10, linetype="dashed", color = "red") 

a

```

Acropora, Montipora, other branching (p.dam, stylo), other hermatypic coral are the dominant reef building taxa. 

Recategorise  cover into these categories:

```{r}
# Nested if else to create a new variable where genera are sorted into benthic category type.

benthic.cats.other <- t.cover.freq %>%
  mutate(Category = 
      ifelse(Genera %in% c("Acanthastrea", "Alveopora","Astrea","Euphyllia","Paragoniastrea","Pectinia","Platygyra", "Plesiastrea","Porites","Psammocora","Cyphastrea","Homophyllia.sp.","Lobophyllia","Turbinaria"),
             "Hard_coral",
      ifelse(Genera %in% c("Acropora_br","Acropora_pl"),
             "Acropora",
      ifelse(Genera %in% c("Pocillopora","Stylophora"),
             "Other_branching",
      ifelse(Genera %in% c("Montipora_enc","Montipora_fol"),
             "Montipora",
      ifelse(Genera %in% c("Amansia","Amphiroa","Asparagopsis","Bryopsis","Caulerpa_cupre","Caulerpa_racemosa", "Chaetomorpha","Chlorodesmis.sp.","Codium.fragile","Colpomennia.sinuosa","Dasycladus","Dictyopteris","Dictyota","Halymenia","Homophyllia.sp", "Hormosira","Hydroclathrus","Hypnea","Laurencia","Padina","Peysonnelia","Plocamium","Sargassum","Ulva","Valonia"), 
             "Macroalgae",
      ifelse(Genera %in% c("Soft_coral","Anemone"), 
             "Soft_coral",
      ifelse(Genera %in% c("Sponge","Zooanthid","Benthic_invert"), 
             "Other_benthic_invert",
      ifelse(Genera %in% c("Black_cyano"),
             "Lyngbya",
      ifelse(Genera %in% c("Turf"),
             "Turf",
      ifelse(Genera %in% c("Turf_sediment"),
             "Turf_sediment_matrix",
      ifelse(Genera %in% c("Cyanofil","Red_cyano"),
             "Cyano_matrix",
      ifelse(Genera %in% c("Sediment"),
             "Sediment",
      ifelse(Genera %in% c("Sand","Rubble","Dead_coral"),
             "Other_substrate", 
      ifelse(Genera %in% c("CCA","Rhodoliths","Halimeda","Tricleocarpa.fragilis"),
             "Other_calcifier","NA"
             )))))))))))))))

View(benthic.cats.other)

str(benthic.cats.other)

benthic.cats.other$Category <- as.factor(benthic.cats.other$Category)


# Relevel Date so that December comes first 
benthic.cats.other$Date <- factor(benthic.cats.other$Date, levels = c("Dec", "April"))

```


```{r}

# Look at the levels
levels(benthic.cats$Category)
head(benthic.cats)

# Filter out Hard_coral
hard_coral <- benthic.cats %>% filter(Category == "Hard_coral")

# Summarise data frame (per coral taxa, per time point)
benthic.cov.sum <- benthic.cats.other %>% group_by(Date,Category,Transect_ID) %>% 
  summarise(rel_mean = mean(freq, na.rm = TRUE))
            #,
            #SD_freq = sd(freq),
            #SE_freq = sd(freq/sqrt(n()))) 

benthic.cov.sum1 <- benthic.cov.sum %>% group_by(Date,Category) %>% 
  summarise(rel_sum = mean(rel_mean, na.rm = TRUE),
            SD_freq = sd(rel_mean),
            SE_freq = sd(rel_mean/sqrt(n()))) 

# plot relative coral cover of genera over time
colours <- c("#61B746","#518794")

# order taxa by grouping for plotting 

benthic.cov.sum1$Category <- factor(benthic.cov.sum1$Category, levels = c("Montipora","Acropora","Other_branching","Hard_coral", "Other_calcifier", "Other_benthic_invert","Soft_coral","Macroalgae","Turf","Turf_sediment_matrix","Lyngbya","Cyano_matrix", "Sediment","Other_substrate"))

plot <- ggplot(benthic.cov.sum1, aes(Category, rel_sum, fill = Date)) +
 geom_col(position = "dodge",stat = "identity", alpha = 0.5) +
 geom_errorbar(aes(ymin = rel_sum - SE_freq, ymax = rel_sum + SE_freq), 
               width=0.2, position = position_dodge(.9), color = "black", size = 1.3) +
  scale_y_continuous(limits = c(0,25), breaks =c(0,5,10,15,20,25)) +
  #geom_label(stat = 'summary', fun.y=mean, aes(label = round(..y.., 1)), nudge_x =0.5 , hjust = 3) +
  xlab ("Benthic group") +
  ylab ("Absolute cover (%)") + 
  ggtitle("Absolute cover over time per benthic group") +
 theme_few() +
  scale_fill_manual(values = colours)

a <- plot + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) #+ #geom_hline(yintercept=10, linetype="dashed", color = "red") 

a

ggplot(benthic.cats.other, aes(x = Date, y = freq, fill = Category)) +
      geom_bar(position="fill", stat="identity") +
      scale_fill_viridis_d() + theme_few()

```


### C) PCA - Generate driver variables that can be used in modelling disease occurence and disease severity.


Data cleaning steps

Make each group its own variable using pivot wider.

```{r}

str(benthic.cov.sum)

# Transpose data-set
# Rows for each category, for 24 transects.

# Use pivot wider to 

wide.benthic.cats <- benthic.cov.sum %>% pivot_wider(names_from = Category, values_from = rel_mean)
str(wide.benthic.cats)
# Add total hard coral onto this and write a csv file. 
head(hc_trana)
hc_tranb <- hc_trana %>% group_by(Date, Transect_ID) %>% summarise(totalhardcoral = mean(t_hardcoral))

benthic.cats.final <- bind_cols(wide.benthic.cats, hc_tranb) %>% select()

# Select out columns of interest 
write.csv(benthic.cats.final, "Benthic_categories.csv")
```

Look for correlations between variables

Correlation matrix

```{r}
str(wide.benthic.cats)
res <- cor(wide.benthic.cats[,c(3:16)])
round(res, 2)

# Visualise correlations
library(corrplot)

plot <- corrplot(res, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)

# Positive correlations are shown in blue 
# Negative correlations in red

```

Looks like we do see correlations between some of these variables.

Try a multivariate statistical technique to reduce the number of variables for use in later analysis, and to account for correlations between variables.

Data cleaning prior to PCA analysis:

```{r}

# Make a new dataframe with with just the variables to be analysed by PCA 

# Ungroup data by date 

benthic.groups.multivar <- wide.benthic.cats %>% ungroup() %>%
  select(3:16)

```

PCA

```{r}

benthicPCA <- princomp(benthic.groups.multivar, cor = FALSE)

summary(benthicPCA)

screeplot(benthicPCA, type = 'lines')

fviz_eig(benthicPCA, addlabels = TRUE, ylim = c(0,50))


# Loadings as a .csv

printl <- loadings(benthicPCA)
threshold <- 0.1
dd <- as.data.frame.matrix(printl)

dd[abs(dd) < threshold] <- NA

# Export matrix as csv file for supplemental materials
write.csv(dd,"Benthic_PCA_loadings.csv")

biplot(benthicPCA)

# Eigenvalues
eig.val <- get_eigenvalue(benthicPCA)
eig.val
  
write.csv(eig.val,"Benthic_PCA_eigenvalues.csv")


# Results for Variables
res.var <- get_pca_var(benthicPCA)
res.var$coord          # Coordinates
res.var$contrib        # Contributions to the PCs
res.var$cos2           # Quality of representation 

# Visualise which variables contribute the most to each dimension
corrplot(res.var$contrib, is.corr=FALSE)   

# Contributions of variables to PC1
fviz_contrib(benthicPCA, choice = "var", axes = 1, top = 10)
# Contributions of variables to PC2
fviz_contrib(benthicPCA, choice = "var", axes = 2, top = 10)
# Contributions of variables to PC3
fviz_contrib(benthicPCA, choice = "var", axes = 3, top = 10)
# Contributions of variables to PC4
fviz_contrib(benthicPCA, choice = "var", axes = 4, top = 10)

# Results for individuals (Eigenvectors)
res.ind <- get_pca_ind(benthicPCA)
eignvec <- res.ind$coord          # Coordinates
res.ind$contrib        # Contributions to the PCs
res.ind$cos2           # Quality of representation 

str(eignvec)

dd.eignvec <- as.data.frame.matrix(eignvec)

metadata <- wide.benthic.cats %>%
  select(1:2)

d.eigenvec <- dd.eignvec %>%
  select(1:4) %>%
  cbind(metadata) 

write.csv(d.eigenvec,"Benthic_PCA_eigenvectors.csv")

```

Visualise results as a loading plot 

```{r}

plottingloads <- read.csv("Benthic_PCA_loadings.csv")
str(plottingloads)

plottingloads$X <- as.factor(plottingloads$X)
plottingloads$Category <- plottingloads$X


# Filter out the first 4 PCs
pl <- plottingloads %>% select(2:5,16)
head(pl)

# Transpose data frame for plotting 

# Transpose the data frame
pl.long <- pivot_longer(data = pl,values_drop_na = TRUE, names_to = "Comp",cols = Comp.1:Comp.4)
pl.long$Category <- factor(pl.long$Category,levels = c("Montipora", "Acropora", "Other_branching", "Sediment","Turf_sediment_matrix","Lyngbya","Cyano_matrix"))


# Plot this with values and labels switched
pl <- ggplot(pl.long, aes(x = Category, y = value, fill = value)) +
  geom_bar(stat = "identity") +
  #scale_fill_continuous() +
  facet_wrap(~Comp,ncol = 4, dir = "h") + 
  scale_fill_viridis_c() +
  theme_few() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ylim(-0.9,0.9) +
  coord_flip() + geom_hline(yintercept=0, color = "grey") 

```


```{r}

library(factoextra)

fviz_pca_biplot(benthicPCA, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )

groups <- as.factor(wide.benthic.cats$Date)

fviz_pca_ind(benthicPCA,
             geom.ind = "point", axes = c(1,2),
             col.ind = groups, # color by groups
             palette = c("#00AFBB",  "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             ellipse.type = "confidence",
             legend.title = "Groups"
            # repel = TRUE
             )

fviz_pca_ind(benthicPCA, axes = c(2, 3),
             geom.ind = "point",
             col.ind = groups, # color by groups
             palette = c("#00AFBB",  "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             ellipse.type = "confidence",
             legend.title = "Groups"
            # repel = TRUE
             )

fviz_pca_ind(benthicPCA, axes = c(1, 3),
             geom.ind = "point",
             col.ind = groups, # color by groups
             palette = c("#00AFBB",  "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             ellipse.type = "confidence",
             legend.title = "Groups"
            # repel = TRUE
             )

```

# Biplot 

```{r}


fviz_pca_biplot(benthicPCA, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
# Colour individuals by group (Date)
# Colour variables by contribution to the principal componant, 

fviz_pca_biplot(benthicPCA, 
                # Individuals
                geom.ind = "point",
                fill.ind = groups, col.ind = "black",
                pointshape = 21, pointsize = 2,
                palette = "jco",
                addEllipses = TRUE,
                # Variables
                #alpha.var ="contrib", 
                col.var = "contrib",
                gradient.cols = "RdYlBu",
                
                legend.title = list(fill = "Date", color = "Contrib"
                                    ) 
                ) + theme_classic()
  #xlim(-50,50) +
  #ylim(-25,25)

fviz_pca_var(benthicPCA,
             col.var = "black",     # Active variables
             col.quanti.sup = "red" # Suppl. quantitative variables
             )
```


Make a data.frame out of original individual loadings 


```{r}
benthic.groups.multivar <- wide.benthic.cats %>% ungroup() %>%
  select(3:16)

library(sjPlot)
library(sjmisc)

# group
bgm <- wide.benthic.cats %>% ungroup() %>%
  select(3:16)

sjp.pca(benthic.groups.multivar, metadata = wide.benthic.cats)

p <- pca(wide.benthic.cats)
plotloadings(p)

library(PCAtools)

```











