---
title: "Transect-level analysis"
output: html_notebook
author: Charlotte Page
---

# Transect-level statistical analysis

This is code for transect-level statistical analysis on disease dynamics. 


1) Summary statistics


2) Disease prevalence (% of surveyed community showing active disease signs)
    a) Build glm (use weights function to specify number of colonies per transect), family = binomial


3) Disease severity (the best variable for this response needs to be assessed)
    a) Step one: Investigate relationships between disease severity and lesion number 
    b) Step two: Build model (lm, average is normally distributed)
    
    
4) Cluster analysis
    a) Step one: Organise data 
    b) Step two: Calculate distance matrix based on severity and prevalence; and each seperately compared to a zero variable. 
   
    
Predictor variables for models in 2) and 3):

+ TP (Dec,April)
+ Site (EB,SB)
+ Total Montipora cover (%)
+ TP*Site interaction

Libraries and data
```{r}
# Libraries 
library(ggplot2)
library(dplyr)
library(tidyr)
library(multcomp)
library(lme4)
library(effects)
library(lmerTest)
library(afex)
library(car)
library(magrittr)
library(ggeffects)
library(sjmisc)
library(splines)
#library(glmmTMB)#
library(DHARMa)
library(devtools)
library(pbkrtest)

library(sjPlot)
library(sjmisc)
library(sjlabelled)

# Transect-level analysis rda file
transect_analysis <- read.csv("transect_level_analysis.csv")

# Make a new variable called transect ID

transect_analysis$Transect_ID <- paste(transect_analysis$TP, transect_analysis$Transect, sep = "_")

str(transect_analysis)

```

### 1) Summary statistics


Summary data for manuscript 

For table (code this into an actual table?)

```{r}


summaryprev <- by(transect_analysis$prev, transect_analysis$TP, summary)

summaryprev_1 <- transect_analysis %>% group_by(TP) %>% 
  summarise(mean_prev = mean(prev), sd_prev = sd(prev), sample_n = n()) 

summaryprev_1

summaryprev

summarysev <- by(transect_analysis$sev, transect_analysis$TP, summary)
summarysev


summarysev_1 <- transect_analysis %>% group_by(TP) %>% 
  summarise(mean_sev = mean(severity), sd_sev = sd(severity), sample_n = n()) 

str()

summarymonti <- summary(transect_analysis$Montipora)
summarymonti

summarythardcoral <- summary(transect_analysis$totalhardcoral)
summarythardcoral

# Caclulate relative cover of monti

relativemonti <- 6.37/33.28

# Calculate increase in sev over time 

increase <- 15.35 - 9.992

percentincrease <- increase/9.992 

```

For manuscript 

```{r}

# Total corals surveyed at each time point 

number <- transect_analysis %>% group_by(TP) %>% summarise(n = sum(total_corals_surveyed))

summprevSEsite <- transect_analysis %>% group_by(TP, Site) %>% summarise(meanprev = mean(prev),
                                                                 SEprev = sd(prev)/sqrt(n()))

```


### 2) Disease prevalence (% of surveyed community showing active disease signs)
    
    
##### a) Build glm (use weights function to specify number of colonies per transect), family = binomial


GLM with Binomial family

Use a weights argument to set the number of trials (i.e. number of surveyed corals per transect), and use the proportion as the response directly.  

Weights arg = number of trails used to generate each proportion. 


```{r}

hist(transect_analysis$Montipora)

mod1 <- glm(freq ~ Site * TP +  Montipora, weights = total_corals_surveyed, family = binomial(), data=transect_analysis )

summary(mod1)

Anova(mod1, type = 3)

tab_model(mod1, dv.labels = "Transect-level: Prevalence Model",   string.ci = "Conf. Int (95%)",
  string.p = "P-Value")


exp(coef(mod1))/(1+exp(coef(mod1)))

plot.dat <- data.frame(Site = transect_analysis$Site,
                       TP = transect_analysis$TP,
                       Montipora = transect_analysis$TP,
                       fit = predict(mod1< Site:TP)) 
#convert those logit values to probabilities
plot.dat$fit_prob <- exp(plot.dat$fit)/(1+exp(plot.dat$fit))

```

Test model assumptions 

```{r}
simulationOutput <- simulateResiduals(fittedModel = mod1, plot = F)
residuals(simulationOutput)

plot(simulationOutput)
```

Looks good! 

See that there is a significant interaction between TP and Site. We can do a posthoc test to find out where this difference is.

Code followed:
https://www.rpubs.com/daharo_calpoly/502695 

```{r}
library(multcomp)

#Investigating the interaction

#Group the data by groups of interest by creating a new variable
transect_analysis$groups<-interaction(transect_analysis$Site, transect_analysis$TP)

#Create a post-hoc model
model_posthoc<-with(data = transect_analysis, glm(freq ~ groups, family=binomial(), weights = total_corals_surveyed))

summary(glht(model_posthoc, linfct = mcp (groups = 
                                            c("(EB.April)-(SB.April)=0",
                                              "(EB.December)-(SB.December)=0",
                                              "(EB.December)-(EB.April)=0",
                                          "(SB.December)-(SB.April)=0"))), test = adjusted("holm"))

prevmonti<- plot_model(mod1, type = "pred", terms = "Montipora[all]", show.data = TRUE) + theme_classic() 
plot_model(mod1, type = "int", terms = "SiteSB:TPDecember", show.values = T) + theme_classic() 

m <- plot_model(mod1, type = "pred", terms = "Montipora", show.data = T, show.values = TRUE) + theme_classic()
m


effect_plot(mod1, pred = Montipora, interval = TRUE, plot.points = TRUE,point.size = 1) + theme_classic()



plot_summs(mod1)

```

### 3) Disease severity 

#####    b) Step two: Build model (lm, average is normally distributed)

```{r}
library(jtools)

hist(transect_analysis$severity)

transect_analysis$TP <- as.factor(transect_analysis$TP)
transect_analysis$Site <- as.factor(transect_analysis$Site)

str(transect_analysis)

mod2 <- lm(severity ~  Site * TP + Montipora, data=transect_analysis)

plot(mod2)
qqnorm(mod2)

summary(mod2) 

anova(mod2)
tab_model(mod2, dv.labels = "Transect-level: Severity Model",   string.ci = "Conf. Int (95%)",
  string.p = "P-Value")

# Generate anova table

library(jtools)

Anova(mod2, type = 3)

sevmonti <-plot_model(mod2, type = "pred", terms = "Montipora", show.data = TRUE) + theme_classic()

plot_model(mod2, type = "int", terms = "SiteSB:TPDecember", show.values = T) + theme_classic() 

m1 <- plot_model(mod2, type = "pred", terms = "Montipora", show.data = T, show.values = TRUE) + theme_classic()
m1

m1 <- plot_model(mod2, type = "pred", terms = "Montipora", show.data = T, show.values = TRUE) + theme_classic()
m1

effect_plot(mod2, pred = Montipora, interval = TRUE, plot.points = TRUE) + theme_classic()

plot_summs(mod2)

```







```{r}

hist(transect_analysis$severity)

hist(log(transect_analysis$severity))

hist(transect_analysis$Montipora)

hist(log(transect_analysis$Montipora))

# Try running an lmer without transforming - it looks pretty nice and normal tbh

mod3 <- lmer(severity ~  Site * TP + Montipora + (1|Transect), data=transect_analysis)

plot(mod3)

summary(mod3)

# This errors - try without the random effects structure (looks like transect represents v little variance = 0)

mod4 <- lm(severity ~  Site * TP + Montipora, data=transect_analysis)

plot(mod4)

summary(mod4)

anova(mod4)

ggplot(transect_analysis, aes(severity,Montipora)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  theme_classic()

# Colour points by TP and add regression coefficients onto plot 

# TP1 - Greens
# Hex code: #5AB350

# TP2 - Blues
# Hex code: 2DB3B0

montisevplot <- ggplot(transect_analysis, aes(Montipora,severity, colour = TP)) +
  geom_point() +
  geom_abline(intercept = 12.371, slope = 0.3612) +
  theme_classic() +
  scale_color_manual(values=c("#5AB350", "#2DB3B0")) +
  theme_classic()
montisevplot

# Plotting predicted values with confidence intervals (see plotting below)


newdat.lme = data.frame(grp = dat$grp,
                        x1 = dat$x1,
                        x2 = median(dat$x2) )
head(newdat.lme)

```
```





Use a parametric bootstrap to test for the effects of variables:

PBmodcomp in the pbkrtest() implements parametric bootstrapping by comparing a full model to a null one. The function simulates data from the null model (the response variable), and then fits these data to the null and full model generated over many simulations to derive a p-value using liklihood testing. 

The advantage of this method is that we do not assume a null dsitribution (i.e. chi-square) but instead create our own null distribution from the data and the model.


```{r}




```






The code in the chunk below is from: 

https://www.r-bloggers.com/2015/02/generating-anova-like-table-from-glmm-using-parametric-bootstrap/


```{r}
anova_merMod<-function(model,rand,w=NULL,seed=round(runif(1,0,100),0),nsim=50){
  data<-model@frame
  if(!is.null(w)){
    data<-data[,-grep("(weights)",names(data))]
  }
  
  resp<-names(model.frame(model))[1]
  #generate a list of reduced model formula
  fs<-list()
  fs[[1]]<-as.formula(paste(resp,"~ 1 +",rand))
  nb_terms<-length(attr(terms(model),"term.labels"))
  if(nb_terms>1){
    for(i in 1:nb_terms){
      tmp<-c(attr(terms(model),"term.labels")[1:i],rand)
      fs[[i+1]]<-reformulate(tmp,response=resp)
    }      
  }
  #fit the reduced model to the data
  
  fam<-family(model)[1]$family
  if(fam=="gaussian")
    {
    m_fit<-lapply(fs,function(x) lmer(x,data,REML=FALSE))
  } else if(fam=="binomial"){
    m_fit<-lapply(fs,function(x) glmer(x,data,family=fam,weights=w))
  }  else {
    m_fit <-lapply(fs,function(x) glm(x,data,family=fam,weights=w))}
  
  #compare nested model with one another and get LRT values (ie increase in the likelihood of the models as parameters are added)
  tab_out<-NULL
  
  for(i in 1:(length(m_fit)-1)){
    comp<-PBmodcomp(m_fit[[i+1]],m_fit[[i]],seed=seed,nsim=nsim)    
    term_added<-attr(terms(m_fit[[i+1]]),"term.labels")[length(attr(terms(m_fit[[i+1]]),"term.labels"))]
    #here are reported the bootstrapped p-values, ie not assuming any parametric distribution like chi-square to the LRT values generated under the null model
    #these p-values represent the number of time the simulated LRT value (under null model) are larger than the observe one
    tmp<-data.frame(term=term_added,LRT=comp$test$stat[1],p_value=comp$test$p.value[2])
    tab_out<-rbind(tab_out,tmp)
    print(paste("Variable ",term_added," tested",sep=""))
  }  
  print(paste("Seed set to:",seed))
  return(tab_out)  
}
```

```{r}

anova_merMod(model = mod1, w = total_corals_surveyed)

```





```{r}

View(transect_level)

mod1 <- glmer(freq ~ Site * TP +  Montipora + (1|Transect), weights = total_corals_surveyed, family = binomial(), data=transect_analysis)

mod2 <- glmer(freq ~ Site * TP + total_corals_surveyed + (1|Transect), weights = total_corals_surveyed, family = binomial(), data=transect_analysis)

summary(mod1)
summary(mod2)


Anova(mod1)
anova(mod1)

library(MuMIn)
r.squaredGLMM(mod1)

```

Model diagnostics
```{r}
simulationOutput <- simulateResiduals(fittedModel = mod1, plot = F)
residuals(simulationOutput)

plot(simulationOutput)
```

Looks good! 

See that there is a significant interaction between TP and Site. We can do a posthoc test to find out where this difference is.

```{r}
library(lsmeans)
lsmeans(mod1, pairwise ~ Site | TP, adjust = "Tukey")
lsmeans(mod1, pairwise ~ TP | Site, adjust = "Tukey")
```

Difference lies in December between EB and SB (p < 0.01)

### 2) Disease severity (mean % of colony)

-	Log transform the Area variable and then try and do a normal linear mixed effects model.

```{r}

hist(transect_analysis$severity)

hist(log(transect_analysis$severity))

hist(transect_analysis$Montipora)

hist(log(transect_analysis$Montipora))

# Try running an lmer without transforming - it looks pretty nice and normal tbh

mod3 <- lmer(severity ~  Site * TP + Montipora + (1|Transect), data=transect_analysis)

plot(mod3)

summary(mod3)

# This errors - try without the random effects structure (looks like transect represents v little variance = 0)

mod4 <- lm(severity ~  Site * TP + Montipora, data=transect_analysis)

plot(mod4)

summary(mod4)

anova(mod4)

ggplot(transect_analysis, aes(severity,Montipora)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  theme_classic()

# Colour points by TP and add regression coefficients onto plot 

# TP1 - Greens
# Hex code: #5AB350

# TP2 - Blues
# Hex code: 2DB3B0

montisevplot <- ggplot(transect_analysis, aes(Montipora,severity, colour = TP)) +
  geom_point() +
  geom_abline(intercept = 12.371, slope = 0.3612) +
  theme_classic() +
  scale_color_manual(values=c("#5AB350", "#2DB3B0")) +
  theme_classic()
montisevplot

# Plotting predicted values with confidence intervals (see plotting below)


newdat.lme = data.frame(grp = dat$grp,
                        x1 = dat$x1,
                        x2 = median(dat$x2) )
head(newdat.lme)

```

# Colours 

Disease - purples 

No disease - greens 

Theme for plotting 

Prevalence plot 

Prevalence by site over time
```{r}

transect_analysis$TP <- factor(transect_analysis$TP, levels = c("December","April"))

# Summarise prevalence values for plotting - with SE!

prev_sum <- transect_analysis %>% 
  group_by(TP, Site) %>% 
  summarise(mean_Prev=mean(prev),
  sd_Prev = sd(prev),
  SE_Prev = sd(prev)/sqrt(n()))


prevplot <- 
  ggplot(data = prev_sum, aes(Site, mean_Prev)) +
 geom_point() +
    geom_errorbar(aes(ymin = mean_Prev - SE_Prev, ymax = mean_Prev + SE_Prev), width=0.1) +
  # geom_point(data = transect_analysis, aes(Site, prev)) +
  scale_y_continuous(limits = c(20,100), breaks = c(20,30,40,50,60,70,80,90,100)) +
 theme_classic() +
  facet_wrap(~TP) +
  theme_classic()

prevplot

prevplot +  geom_point(transect_analysis, aes(Site,prev))

# Colour 

ggplot(transect_analysis)
```

Severity plots 

Severity over time 
```{r}

# Summarise severity values for plotting - with SE!
sev_sum <- transect_analysis %>% 
  group_by(TP,Site) %>% 
  summarise(mean_Prev=mean(severity),
  sd_Prev = sd(severity),
  SE_Prev = sd(severity)/sqrt(n()))


sevplot <- ggplot(sev_sum, aes(Site, mean_Prev)) +
 geom_point() +
 geom_errorbar(aes(ymin = mean_Prev - SE_Prev, ymax = mean_Prev + SE_Prev), width=0.1) +
  scale_y_continuous(limits = c(0,20), breaks = c(0,5,10,15,20)) +
 theme_classic() +
 facet_wrap(~TP) 

sevplot


```

Severity x montipora plot  

Bring plots together 

###  Plotting results

```{r}
grid <- plot_grid(prevplot,sevplot,montisevplot, labels = c('A', 'B','C'), label_size = 12)

grid
```

## B. Cluster analysis

### Plotting results





```{r}
summ_december <- transect_analysis %>% filter(TP == "December") %>%
  mutate(zero = "0")

cluster_dec <- summ_december[,c(7,9)]
cluster_dec.sev <- summ_december[,c(7,27)]
cluster_dec.prev <- summ_december[,c(9,27)]

summ_april <- transect_analysis %>% filter(TP == "April") %>%
  mutate(zero = "0")

cluster_april <- summ_april[,c(7,9)]
cluster_april.sev <- summ_april[,c(7,27)]
cluster_april.prev <- summ_april[,c(9,27)]

```



```{r}

# DECEMBER

rownames(cluster_dec) <- summ_december$Transect
rownames(cluster_dec.sev) <- summ_december$Transect
rownames(cluster_dec.prev) <- summ_december$Transect

# Create a matrix - similarity coefficient = Euclidean distance 

Dec.sim <- dist(cluster_dec, method = "euclidean")
Dec.sim.sev <- dist(cluster_dec.sev, method = "euclidean")
Dec.sim.prev <- dist(cluster_dec.prev, method = "euclidean")

Dec.sim.cluster <- hclust(Dec.sim, method = "ward")
Dec.sim.cluster.sev <- hclust(Dec.sim.sev, method = "ward")
Dec.sim.cluster.prev <- hclust(Dec.sim.prev, method = "ward")

plot(Dec.sim.cluster)
plot(Dec.sim.cluster.sev)
plot(Dec.sim.cluster.prev)

# convert it to a dendogram 

plot(as.dendrogram(Dec.sim.cluster),ylab = "Euclidean distance")
b <- plot(as.dendrogram(Dec.sim.cluster.sev),ylab = "Euclidean distance")
c <- plot(as.dendrogram(Dec.sim.cluster.prev),ylab = "Euclidean distance")


# APRIL

rownames(cluster_april) <- summ_april$Transect
rownames(cluster_april.sev) <- summ_april$Transect
rownames(cluster_april.prev) <- summ_april$Transect

# Create a matrix - similarity coefficient = Euclidean distance 

April.sim <- dist(cluster_april, method = "euclidean")
April.sim.sev <- dist(cluster_april.sev, method = "euclidean")
April.sim.prev <- dist(cluster_april.prev, method = "euclidean")

April.sim.cluster <- hclust(April.sim, method = "ward")
April.sim.cluster.sev <- hclust(April.sim.sev, method = "ward")
April.sim.cluster.prev <- hclust(April.sim.prev, method = "ward")

plot(April.sim.cluster)
plot(April.sim.cluster.sev)
plot(April.sim.cluster.prev)

# convert it to a dendogram 

plot(as.dendrogram(April.sim.cluster),ylab = "Euclidean distance")
plot(as.dendrogram(April.sim.cluster.sev),ylab = "Euclidean distance")
plot(as.dendrogram(April.sim.cluster.prev),ylab = "Euclidean distance")
```







