---
title: "Colony-level analysis"
output: html_notebook
author: Charlotte Page
---

# Colony-level statistical analysis

This is code for colony-level statistical analysis on disease dynamics. 

**Notes**

Model-selection using lasso - Lasso reduces the variable selection problem to the selection of a single penalty parameter and then the coefficient estimation follows naturally. The key parameter when fitting the lasso is λ, the penalty parameter. We choose λ based on an iterative testing approach, where BIC (lowest) is used to choose lambda.

1) Disease occurance (presence/absence)
    a) GLMM specification 
    b) Lasso: cross validation to select lambda
    c) Model checking
    d) Results

2) Disease severity (% of colony impacted by disease)
    a) GLMM specification 
    b) Lasso: cross validation to select lambda  
    c) Model checking
    d) Results

```{r}
# Libraries 
library(ggplot2)
library(dplyr)
library(tidyr)
library(multcomp)
library(lme4)
library(effects)
library(lmerTest)
library(afex)
library(car)
library(magrittr)
library(ggeffects)
library(sjmisc)
library(splines)
#library(glmmTMB)#
library(DHARMa)
library(devtools)
library(pbkrtest)
library(glmmLasso)
library(sjPlot)
library(sjmisc)
library(sjlabelled)

# Colony-level analysis rda file

cl <- read.csv("Colony_data_analysis.csv")
load("Colony_data_analysis.rda")
str(disease.vars)

# Make site variable

disease.vars$Site <- ifelse(grepl("EB",disease.vars$Transect_ID),"EB","SB")

disease.vars$Site <- as.factor(disease.vars$Site)
str(disease.vars)


head(disease.vars)



```
    
1) Disease occurance (presence/absence)

1) Disease occurance (presence/absence)
    a) Lasso: cross validation to select lambda
    b) Run glmmLASSO Model 
    c) Run glmer for visualisation of results and diagnostics (this may not be nec, see if we can get glmmLASSO to give fixed effects)
    d) Results 
    
    # Note: SEs are conditional on model-selection techniques.
    
```{r}


str(disease.vars)

library(glmmLasso)
library(dplyr)
library(tidyr)

# Try using other code. 

#create a list of lambda’s to try. 

# First look to see which range of lambda values have the lowest BIC
Lambda_try <- seq(100,0,by=-5)

# Explore this further 
Lambda_try <- c(20:30)

#initialise a vector to store the BIC’s
BIC_from_Lambda_try_models <- c()

# loop through the Lambda_try to see what the BIC of each resulting model is. 

for (i in 1: length(Lambda_try)){
  
#customise formula, random effect, family and dataset
glm1 <- glmmLasso(occ ~ 
                    Site + 
                    TP + 
                    Growth_form + 
                    Colour_morph + 
                    Size + 
                    Colour + 
                    new_Other_mortality + 
                    new_Dim.1 + 
                    new_Dim.2 + 
                    new_Dim.3 + 
                    new_Dim.4 +
                    Site:TP + 
                    Growth_form:TP +
                    Colour_morph:TP + 
                    Colour:TP + 
                    Size:TP +
                    new_Other_mortality:TP + 
                    Growth_form:Site +
                    Colour_morph:Site + 
                    Colour:Site + 
                    Size:Site +
                    new_Other_mortality:Site +
                    new_Dim.1:Site + 
                    new_Dim.2:Site + 
                    new_Dim.3:Site + 
                    new_Dim.4:Site,
          
                    rnd = list(Transect_ID=~1), family = binomial(), data = disease.vars, lambda = Lambda_try[i])

BIC_from_Lambda_try_models[i] <- glm1$bic

}

#plot the BIC’s versus Lambda_try

plot(Lambda_try, BIC_from_Lambda_try_models)



```
  b) Run glmmLASSO Model 

```{r}

str(disease.vars)
library(lmerTest)

glm2 <- glmmLasso(occ ~ 
                    Site + 
                    TP + 
                    Growth_form + 
                    Colour_morph + 
                    Size + 
                    Colour + 
                    new_Other_mortality + 
                    new_Dim.1 + 
                    new_Dim.2 + 
                    new_Dim.3 + 
                    new_Dim.4 +
                    Site:TP + 
                    Growth_form:TP +
                    Colour_morph:TP + 
                    Colour:TP + 
                    Size:TP +
                    new_Other_mortality:TP + 
                    Growth_form:Site +
                    Colour_morph:Site + 
                    Colour:Site + 
                    Size:Site +
                    new_Other_mortality:Site +
                    new_Dim.1:Site + 
                    new_Dim.2:Site + 
                    new_Dim.3:Site + 
                    new_Dim.4:Site,
          
                   rnd = list(Transect_ID=~1), family = binomial(), data = disease.vars, lambda=23,final.re=TRUE)

# Model summary 
summary(glm2)

glm2$coefficients
glm2$ranef
plot(glm2$fitted.values)
summary(glm2)

## odds ratios only

exp(coef(glm2))


# Need to have a play around here and see if we can get fixed effects in interactions to also have a coeficient. 




library(ggeffects)

# dat is a data frame with marginal effects

library(sjPlot)
library(sjlabelled)
library(sjmisc)
library(ggplot2)
effects::allEffects(glm1)

theme_set(theme_sjplot())

plot(glm1)

tibble(
  variables  = glm1 %>% tidy() %>% select(glm1),
  log_OR                 = coef(glm1)) %>% 
  cbind(odds.ratio(glm1)) %>% 
  mutate(percent_change  = ifelse(OR < 1, (1/OR - 1)*-100, (OR - 1)*100  ),
         lower_percent_change = ifelse(lower_OR < 1, (1/lower_OR - 1)*-100, (lower_OR - 1)*100  ),
         upper_percent_change = ifelse(upper_OR < 1, (1/upper_OR - 1)*-100, (upper_OR - 1)*100  )) %>%
  mutate_if(is.numeric, ~round(., 3)) %>% 
  select(variables, log_OR, OR, percent_change()) # %>% 
  #kable()

library(effects)

result <- allEffects(glm1$coefficients)

```
 
c) Run glmer for visualisation of results and diagnostics
 
```{r}

library(DHARMa)

glmm.occ.full <- glmer(occ ~ 
                    Site + 
                    TP + 
                    Growth_form + 
                    Colour_morph + 
                    Size + 
                    Colour + 
                    new_Other_mortality + 
                    new_Dim.1 + 
                    new_Dim.2 + 
                    new_Dim.3 + 
                    new_Dim.4 +
                    Site:TP + 
                    Growth_form:TP +
                    Colour_morph:TP + 
                    Colour:TP + 
                    Size:TP +
                    new_Other_mortality:TP + 
                    Growth_form:Site +
                    Colour_morph:Site + 
                    Colour:Site + 
                    Size:Site +
                    new_Other_mortality:Site +
                    new_Dim.1:Site + 
                    new_Dim.2:Site + 
                    new_Dim.3:Site + 
                    new_Dim.4:Site
           + (1|Transect_ID),
                    family = binomial(), data = disease.vars)

# Check diagnostics in DHARMa
plot(simulateResiduals(glmm.occ.full))

glmm.occ.reduced <- glmer(occ ~ 
                    Site + 
                    TP + 
                    Size +
                    new_Dim.2 + 
                    Size:TP +
                    Size:Site +
                    new_Dim.2:Site + 
           + (1|Transect_ID),
                    family = binomial(), data = disease.vars)

plot(simulateResiduals(glmm.occ.reduced))

```


```{r}

library(effects)

summary(glmm.occ.reduced)
plot(allEffects(glmm.occ.reduced))

# Get transformed table of coefficients

allEffects(glmm.occ.reduced) %>% summary()


# Make a forest plot of selected variables and model coefficients

plot_model(glmm.occ.reduced, type = "std") +theme_538()

pmo <- plot_model(glmm.occ.reduced, 
           transform      = NULL, 
           show.intercept = T,
           show.values    = TRUE, 
           value.offset   = .3) +theme_minimal()


# eemeans try and plot interaction 
library(emmeans)
emmip(glmm.occ.reduced, Size ~ TP, CIs=F, plotit=T)+theme_classic()

b <- emmip(glmm.occ.reduced, TP ~ Size, type = "response") + theme_minimal()
c <- emmip(glmm.occ.reduced, Site ~ Size, type = "response") + theme_minimal()

# Eek do you really want to show CI here...I think NOT 
# Probably dont add to the final model (points, CI)
#plot_model(glmm.occ.reduced, type = "pred", terms = #c("new_Dim.2","Site"),show.data = T)+theme_blank()

# Replot 

d <- plot_model(glmm.occ.reduced, type = "pred", terms = c("new_Dim.2","Site"))+ theme_minimal()



library(patchwork)
ok <- (pmo + b+c+d)


```

REPORT RESULTS OF GRAPHS (QUOTE PROBABILITIES _ MAKE A TABLE < OR ADD TO PLOT
MAKE A DISCLAIMER THAT SE WERE NOT USED BEYOND INITIAL PLOTTING OF REGRESSION MODEL RESULTS _ INSTEAD WE DISCUSS RESULTS BASED ON COEFFCIENTS AND MODELLED PREDICTED VALUES)

https://yury-zablotski.netlify.app/post/multiple-logistic-regression-with-interactions/ (ADD NUMBERS TO THIS)

REPEAT THIS FOR AREA (SEVERITY)

c) Interpretation 

Calculate

Plot the coefficients

```{r}
# https://www.mihiretukebede.com/posts/2020-09-30-2020-09-30-plotting-model-coefficients-in-a-forest-plot/
# Coefficient plot 
occ.coef <- as.data.frame(glm1$coefficients)

coef_occ <- occ.coef %>% mutate(variable=rownames(occ.coef))


plot_glm <- ggplot(coef_occ, aes(x=variable, y=glm1$coefficients)) +
  geom_point(shape = 15,
             size  = 4, width = 0.1,
             position = "dodge", color="black") + 
  theme(axis.title = element_text(face = "bold")) +
  xlab("Variables") + ylab("Beta coeffecients GLMM Lasso") +
  coord_flip() + 
  geom_hline(yintercept = 0, color = "red", size = 1) +
   theme(axis.title = element_text(size = 17)) + 
  theme(axis.text = element_text(size = 14)) 
plot_glm

m4_df <- coef(summary(glm1)) %>%
    data.frame() %>%
    tibble::rownames_to_column("term") %>%
    rename(estimate = Estimate, std.error = StdErr)
m4_df

dwplot(m4_df)
plot_model(m4_df)

# Coefficient plot 
occ.coef.ran <- as.data.frame(glm1$ranef)

coef_occ.ran <- occ.coef.ran %>% mutate(variable=rownames(occ.coef.ran))

plot_glm_random <- ggplot(coef_occ.ran, aes(x=variable, y=glm1$ranef)) +
  geom_point(shape = 15,
             size  = 4, width = 0.1,
             position = "dodge", color="black") + 
  theme(axis.title = element_text(face = "bold")) +
  xlab("Variables") + ylab("Beta coeffecients GLMM Lasso") +
  coord_flip() + 
  geom_hline(yintercept = 0, color = "red", size = 1) +
   theme(axis.title = element_text(size = 17)) + 
  theme(axis.text = element_text(size = 14)) +
theme_classic()
plot_glm_random

```

 2) Disease severity (% of colony impacted by disease)
    a) GLMM specification 
    b) Lasso: cross validation to select lambda
 
 
```{r}

hist(disease.vars$Area)
hist(log(disease.vars$Area))



mean(disease.vars$Area)

```
 
```{r}

# Code a dummy variable for healthy/diseased individuals 

# Linear mixed model LASSO 

str(disease.vars)

# Filter out only diseased corals 

disease.vars.sev <- disease.vars %>% filter(Healthy == "Diseased")
disease.vars.sev$Transect_ID <- as.factor(disease.vars.sev$Transect_ID)

library(glmmLasso)
library(dplyr)
library(tidyr)

# Try using other code. 

#create a list of lambda’s to try. 


Lambda_try_sev <- seq(100,0,by=-5)

Lambda_try_sev<-c(30:39,40:50)

#initialise a vector to store the BIC’s
BIC_from_Lambda_try_models_sev <- c()

# loop through the Lambda_try to see what the BIC of each resulting model is. 

for (i in 1: length(Lambda_try_sev)){
  
#customise formula, random effect, family and dataset
glm3 <- glmmLasso(log(Area) ~ Site + 
                    TP + 
                    Growth_form + 
                    Colour_morph + 
                    Size + 
                    Colour + 
                    new_Other_mortality + 
                    new_Dim.1 + 
                    new_Dim.2 + 
                    new_Dim.3 + 
                    new_Dim.4 +
                    Site:TP + 
                    Growth_form:TP +
                    Colour_morph:TP + 
                    Colour:TP + 
                    Size:TP +
                    new_Other_mortality:TP + 
                    Growth_form:Site +
                    Colour_morph:Site + 
                    Colour:Site + 
                    Size:Site +
                    new_Other_mortality:Site +
                    new_Dim.1:Site + 
                    new_Dim.2:Site + 
                    new_Dim.3:Site + 
                    new_Dim.4:Site, rnd = list(Transect_ID=~1), data = disease.vars.sev, lambda = Lambda_try_sev[i])

BIC_from_Lambda_try_models_sev[i] <- glm3$bic

}

#plot the BIC’s versus Lambda_try

plot(Lambda_try_sev,BIC_from_Lambda_try_models_sev)

# 45!
```


```{r}

glm4 <- glmmLasso(log(Area) ~ Site + 
                    TP + 
                    Growth_form + 
                    Colour_morph + 
                    Size + 
                    Colour + 
                    new_Other_mortality + 
                    new_Dim.1 + 
                    new_Dim.2 + 
                    new_Dim.3 + 
                    new_Dim.4 +
                    Site:TP + 
                    Growth_form:TP +
                    Colour_morph:TP + 
                    Colour:TP + 
                    Size:TP +
                    new_Other_mortality:TP + 
                    Growth_form:Site +
                    Colour_morph:Site + 
                    Colour:Site + 
                    Size:Site +
                    new_Other_mortality:Site +
                    new_Dim.1:Site + 
                    new_Dim.2:Site + 
                    new_Dim.3:Site + 
                    new_Dim.4:Site, rnd = list(Transect_ID=~1),
                    data = disease.vars.sev, lambda=45,final.re=TRUE)
glm4$coefficients
str(coef)
coefsev <- as.tibble(coef)

glm4$ranef
summary(glm4)

```

# Model diagnostics 
```{r}

# Full model 

lmm5 <- lmer(log(Area) ~ Site + 
                    TP + 
                    Growth_form + 
                    Colour_morph + 
                    Size + 
                    Colour + 
                    new_Other_mortality + 
                    new_Dim.1 + 
                    new_Dim.2 + 
                    new_Dim.3 + 
                    new_Dim.4 +
                    Site:TP + 
                    Growth_form:TP +
                    Colour_morph:TP + 
                    Colour:TP + 
                    Size:TP +
                    new_Other_mortality:TP + 
                    Growth_form:Site +
                    Colour_morph:Site + 
                    Colour:Site + 
                    Size:Site +
                    new_Other_mortality:Site +
                    new_Dim.1:Site + 
                    new_Dim.2:Site + 
                    new_Dim.3:Site + 
                    new_Dim.4:Site
                    + (1|Transect_ID), data = disease.vars.sev)

# Check diagnostics
plot(simulateResiduals(lmm5))

# Reduced model 

lmm6 <- lmer(log(Area) ~  Site +
               new_Dim.2 + 
               Site:new_Dim.2 +
               (1|Transect_ID), data = disease.vars.sev)

# Check diagnostics
plot(simulateResiduals(lmm6))

```

# Summary of modelling and plotting for final figures 

```{r}

summary(lmm6
        )

# Get transformed table of coefficients

allEffects(lmm6) %>% summary()


# Make a forest plot of selected variables and model coefficients

plot_model(lmm6, type = "std") +theme_538()

pms <- plot_model(lmm6, 
           transform      = NULL, 
           show.intercept = T,
           show.values    = TRUE, 
           value.offset   = .3) +theme_minimal()

pms
f <- plot_model(lmm6, type = "pred", terms = c("new_Dim.2","Site"))+ theme_minimal()

f

library(patchwork)
ok <- (pmo + b+c+d)


```






```{r}

occ.coef <- as.data.frame(glm4$coefficients)

coef_occ <- occ.coef %>% mutate(variable=rownames(occ.coef))


plot_lm <- ggplot(coef_occ, aes(x=variable, y=glm4$coefficients)) +
  geom_point(shape = 15,
             size  = 4, width = 0.1,
             position = "dodge", color="black") + 
  theme(axis.title = element_text(face = "bold")) +
  xlab("Variables") + ylab("Beta coeffecients GLMM Lasso") +
  coord_flip() + 
  geom_hline(yintercept = 0, color = "red", size = 1) +
   theme(axis.title = element_text(size = 17)) + 
  theme(axis.text = element_text(size = 14)) 
plot_lm

plot_model(glmm.occ.reduced, type = "std") + theme_classic() 

# Coefficient plot 
occ.coef.ran <- as.data.frame(glm4$ranef)

coef_occ.ran <- occ.coef.ran %>% mutate(variable=rownames(occ.coef.ran))

plot_glm_random <- ggplot(coef_occ.ran, aes(x=variable, y=glm4$ranef)) +
  geom_point(shape = 15,
             size  = 4, width = 0.1,
             position = "dodge", color="black") + 
  theme(axis.title = element_text(face = "bold")) +
  xlab("Variables") + ylab("Beta coeffecients GLMM Lasso") +
  coord_flip() + 
  geom_hline(yintercept = 0, color = "red", size = 1) +
   theme(axis.title = element_text(size = 17)) + 
  theme(axis.text = element_text(size = 14)) +
theme_classic()
plot_glm_random

```







 
 
 
 
 PLOTTING 
 
 https://stats.stackexchange.com/questions/47020/plotting-logistic-regression-interaction-categorical-in-r
 
 https://blogs.uoregon.edu/rclub/2016/04/05/plotting-your-logistic-regression-models/
 
 https://stackoverflow.com/questions/35366499/ggplot2-how-to-combine-histogram-rug-plot-and-logistic-regression-prediction
 
 Lambda selection 
 
```{r}
#set lambdas... go from 0 to 10^5, in 10 log steps
lambda <- 10^seq(-3,5, length=10)

#dummy vectors of model fit values for each lambda: BIC, AIC, prediction error
 

BIC_vec <- rep(Inf, length(lambda))
AIC_vec <- rep(Inf, length(lambda))


j<-1
for (j in 1:length(BIC_vec)){
 print(paste("Iteration ", j, sep=""))

glm1 <- try(
 glmmLasso(occ ~ Transect + 
             TP + 
             Growth_form + 
             Colour_morph + 
             Size + 
             Colour + 
             Other_mortality + 
             Dim.1 +
             Dim.2 + 
             Dim.3 + 
             Dim.4,
           data = disease.vars, rnd = list(Transect=~1), family = binomial,  lambda = lambda[j],))
   
# code to make it continue anyway if an error occurs
if(class(glm1)!="try-error")
# 
 #save BIC, AIC
 BIC_vec[j]<-glm1$bic
 AIC_vec[j]<-glm1$aic 
 
}
 
```
 
    
    
    
    
    
    
    